---
title: "CMS Synthetic Data (DE-SynPUF) — Chronic Condition & Cost Analysis"
author: "Ann Raymond"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
options(scipen = 999)
```

## 1. Overview

This notebook picks up where **`cms_synpuf_cleaning.Rmd`** leaves off: it
reads four of the five cleaned CSVs that notebook saves to `Data/Clean/`
(Beneficiary, Inpatient, Outpatient, and PDE; Carrier isn't used here) and does
three pieces of downstream analysis on them —

1. **Chronic Condition Prevalence & Correlations** — how common each of the
   11 CMS-flagged chronic conditions is, and a heatmap of their pairwise
   correlations.
2. **Claim Costs by Chronic Condition** — total claim amounts across all
   payers for Inpatient, Outpatient, and Prescription Drug Event claims,
   among beneficiaries with each condition (joined to the Beneficiary
   file's condition flags).
3. **Reconciling Beneficiary Annual Amounts Against Claims** — whether the
   Beneficiary file's annual reimbursement totals tie out against what's
   actually in the Inpatient/Outpatient claims files, using CMS's own
   documented derivation formulas.

It does not re-pull or re-clean anything — if the CSVs in `Data/Clean/`
aren't there yet (or are stale), run `cms_synpuf_cleaning.Rmd` first.

## 2. Setup & load cleaned data

```{r packages}
required_pkgs <- c("tidyverse", "lubridate", "data.table")

installed <- rownames(installed.packages())
missing <- setdiff(required_pkgs, installed)
if (length(missing) > 0) {
  install.packages(missing, repos = "https://cloud.r-project.org")
}

library(tidyverse)
library(lubridate)
library(data.table)
```

Beneficiary and Inpatient are small enough to read whole with `readr`.
Outpatient (75 columns, 780K rows) and PDE (8 columns, 5.6M rows) are read
with `data.table::fread()` and `select =` to pull only the columns this
notebook actually uses — no reason to load 70 unused diagnosis/procedure
columns just to compute a cost summary.

```{r load-data}
beneficiary_all <- read_csv("../Data/Clean/beneficiary_2008_2010.csv", show_col_types = FALSE)

inpatient_clean <- read_csv("../Data/Clean/inpatient_claims_sample_1.csv", show_col_types = FALSE)

outpatient_clean <- fread(
  "../Data/Clean/outpatient_claims_sample_1.csv",
  select = c("beneficiary_id", "claim_end_date", "medicare_payment_amt",
             "beneficiary_blood_deductible_amt", "beneficiary_part_b_deductible_amt",
             "beneficiary_part_b_coinsurance_amt", "primary_payer_paid_amt", "total_claim_amt")
) %>%
  as_tibble() %>%
  mutate(claim_end_date = ymd(claim_end_date))

pde_clean <- fread(
  "../Data/Clean/prescription_drug_events_sample_1.csv",
  select = c("beneficiary_id", "gross_drug_cost_amt")
) %>%
  as_tibble()

cat("beneficiary_all:", nrow(beneficiary_all), "rows x", ncol(beneficiary_all), "cols\n")
cat("inpatient_clean:", nrow(inpatient_clean), "rows x", ncol(inpatient_clean), "cols\n")
cat("outpatient_clean (narrowed):", nrow(outpatient_clean), "rows x", ncol(outpatient_clean), "cols\n")
cat("pde_clean (narrowed):", nrow(pde_clean), "rows x", ncol(pde_clean), "cols\n")
```

## 3. Chronic Condition Prevalence & Correlations

`chronic_condition_count` (computed in the cleaning notebook) is a simple
row-sum across the 11 `chronic_*` flags. This section shows how common each
condition is, then *which* conditions tend to occur together. It collapses to one row per beneficiary
first (a condition counts as present if it was ever flagged TRUE in 2008,
2009, *or* 2010) — otherwise a chronic condition that persists across all
three years would be counted three times for no real reason.

```{r condition-setup}
condition_cols <- c("chronic_alzheimer", "chronic_heart_failure", "chronic_kidney_disease",
                     "chronic_cancer", "chronic_copd", "chronic_depression", "chronic_diabetes",
                     "chronic_ischemic_heart_disease", "chronic_osteoporosis",
                     "chronic_rheumatoid_or_osteoarthritis", "chronic_stroke_tia")

bene_conditions_ever <- beneficiary_all %>%
  group_by(beneficiary_id) %>%
  summarise(across(all_of(condition_cols), any), .groups = "drop")

cat("Unique beneficiaries across 3 years:", nrow(bene_conditions_ever), "\n")
```

### Prevalence

How many beneficiaries have each condition, and what share of all
beneficiaries that is — a sense of scale before looking at how conditions
relate to each other.

```{r condition-prevalence}
tibble(
  condition = condition_cols,
  n_beneficiaries = colSums(bene_conditions_ever[condition_cols]),
  pct_of_beneficiaries = n_beneficiaries / nrow(bene_conditions_ever) * 100
) %>%
  arrange(desc(n_beneficiaries)) %>%
  knitr::kable(digits = 1, format.args = list(big.mark = ","),
               caption = sprintf("Prevalence of each chronic condition (ever flagged in 2008-2010), out of %s beneficiaries",
                                 format(nrow(bene_conditions_ever), big.mark = ",")))
```

### Pairwise correlations

All 11 inputs are 0/1, so a Pearson correlation here is the phi coefficient
— a standard, well-defined measure of association for binary variables.

```{r condition-corr}
cond_matrix <- bene_conditions_ever %>%
  select(all_of(condition_cols)) %>%
  mutate(across(everything(), as.numeric)) %>%
  as.matrix()

cor_matrix <- cor(cond_matrix)
```

```{r condition-corr-heatmap, fig.width=8.5, fig.height=7}
cor_long <- cor_matrix %>%
  as.data.frame() %>%
  rownames_to_column("condition_1") %>%
  pivot_longer(-condition_1, names_to = "condition_2", values_to = "correlation")

ggplot(cor_long, aes(condition_1, condition_2, fill = correlation)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(correlation, 2)), size = 3) +
  scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B", midpoint = 0) +
  theme_minimal(base_size = 10) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), axis.title = element_blank()) +
  labs(title = "Chronic condition correlation matrix")
```

Worth flagging: CMS's own FAQ says disclosure-limitation synthesis distorts
these statistics —
*"much of the interdependence and co-variation among variables have been
altered... univariate and multivariate statistics, such as conditional
correlations and regression coefficients, would not be reliable."* Treat
the prevalence figures and correlations above as a demonstration of the
method, not as real clinical prevalence or comorbidity patterns (see Section 5
for the same caveat applied to dollar amounts).

## 4. Claim Costs by Chronic Condition

This section measures the **total amount of each claim across all payers**
(Medicare, primary payer, and beneficiary responsibility), for beneficiaries
who carry each condition. Inpatient, Outpatient, and PDE claims are joined to
the condition flags from Section 3 (ever flagged). The amounts are for *all*
of a beneficiary's claims, not only claims attributable to the condition
itself.

A beneficiary with 3 conditions contributes their claims to all 3 disease
rows below — that's deliberate (they are flagged with every one of those
conditions), but it means claims are double- or triple-counted across
condition rows, and the per-disease numbers are not independent samples of
each other.

**How each claim amount is built**, per Codebook Appendix 2 ("List of
Reimbursement Variables"):

| Claim type | Total claim amount (all payers) |
|---|---|
| Inpatient | `total_claim_amt` = `medicare_payment_amt` + `utilization_day_count` × `pass_through_per_diem_amt` + `primary_payer_paid_amt` + Part A deductible + Part A coinsurance + blood deductible |
| Outpatient | `total_claim_amt` = `medicare_payment_amt` + `primary_payer_paid_amt` + blood deductible + Part B deductible + Part B coinsurance |
| PDE | `gross_drug_cost_amt` (ingredient cost + dispensing fee + sales tax + vaccine administration fee) |

Inpatient includes the per-diem term because it is part of Medicare's
payment on those claims. Blank deductibles are treated as \$0 in
`total_claim_amt`, so all inpatient claims are included. PDE's gross drug cost is already the
total drug cost regardless of who pays, so the patient-pay amount is part of
it and isn't added again.

```{r claim-amounts}
inpatient_claim_amts <- inpatient_clean %>%
  select(beneficiary_id, claim_amt = total_claim_amt)

outpatient_claim_amts <- outpatient_clean %>%
  select(beneficiary_id, claim_amt = total_claim_amt)

pde_claim_amts <- pde_clean %>%
  transmute(beneficiary_id, claim_amt = as.numeric(gross_drug_cost_amt))
```

```{r condition-flags-long}
condition_flags_long <- bene_conditions_ever %>%
  pivot_longer(all_of(condition_cols), names_to = "condition", values_to = "has_condition") %>%
  filter(has_condition) %>%
  select(beneficiary_id, condition)

```

```{r cost-by-disease-ip-op}
cost_ip <- condition_flags_long %>%
  inner_join(inpatient_claim_amts, by = "beneficiary_id", relationship = "many-to-many") %>%
  group_by(condition) %>%
  summarise(avg_inpatient_claim_amt = mean(claim_amt, na.rm = TRUE),
            median_inpatient_claim_amt = median(claim_amt, na.rm = TRUE),
            n_inpatient_claims = sum(!is.na(claim_amt)), .groups = "drop")

cost_op <- condition_flags_long %>%
  inner_join(outpatient_claim_amts, by = "beneficiary_id", relationship = "many-to-many") %>%
  group_by(condition) %>%
  summarise(avg_outpatient_claim_amt = mean(claim_amt, na.rm = TRUE),
            median_outpatient_claim_amt = median(claim_amt, na.rm = TRUE),
            n_outpatient_claims = sum(!is.na(claim_amt)), .groups = "drop")
```

PDE is 5.6M rows, and joining it fans out further still (a beneficiary's
prescriptions × however many conditions they carry), so this one uses
`data.table` the same way Carrier Claims does in the cleaning notebook, to
keep it fast and light on memory:

```{r cost-by-disease-pde}
cost_pde <- merge(
  as.data.table(condition_flags_long),
  as.data.table(pde_claim_amts),
  by = "beneficiary_id", allow.cartesian = TRUE
)[, .(avg_prescription_claim_amt = mean(claim_amt, na.rm = TRUE),
      median_prescription_claim_amt = median(claim_amt, na.rm = TRUE),
      n_prescriptions = sum(!is.na(claim_amt))), by = condition] %>%
  as_tibble()
```

```{r cost-by-disease-combine}
cost_by_disease <- cost_ip %>%
  full_join(cost_op, by = "condition") %>%
  full_join(cost_pde, by = "condition") %>%
  arrange(desc(avg_inpatient_claim_amt))

cost_by_disease %>%
  knitr::kable(digits = 2, format.args = list(big.mark = ","),
               caption = "Average & median total claim amount (all payers) per claim, among beneficiaries with each chronic condition")
```

## 5. Reconciling Beneficiary Annual Amounts Against Claims

Do `annual_medicare_reimb_ip` / `annual_beneficiary_resp_ip` /
`annual_primary_payer_reimb_ip` (and the OP equivalents) in the Beneficiary
file equal what you get by summing the matching claim-level amounts for
that beneficiary-year? The Codebook's **Appendix 2 ("List of Reimbursement
Variables")** spells out CMS's exact derivation formulas, so this check
uses those instead of a naive `sum(medicare_payment_amt)`:

| Beneficiary variable | CMS's stated derivation (Codebook Appendix 2) |
|---|---|
| `annual_medicare_reimb_ip` | `medicare_payment_amt + utilization_day_count * pass_through_per_diem_amt` |
| `annual_beneficiary_resp_ip` | `beneficiary_deductible_amt + beneficiary_part_a_coinsurance_amt + beneficiary_blood_deductible_amt` |
| `annual_primary_payer_reimb_ip` | `primary_payer_paid_amt` |
| `annual_medicare_reimb_op` | `medicare_payment_amt` |
| `annual_beneficiary_resp_op` | `beneficiary_blood_deductible_amt + beneficiary_part_b_deductible_amt + beneficiary_part_b_coinsurance_amt` |
| `annual_primary_payer_reimb_op` | `primary_payer_paid_amt` |

(Carrier's equivalents are also in Appendix 2, but conditional on
`LINE_PRCSG_IND_CD` — a per-line processing-indicator code dropped during
the Carrier collapse in the cleaning notebook. Reconciling Carrier would
mean re-deriving it from the raw files, which is out of scope here —
matching what was actually asked: Inpatient and Outpatient only.)

Claims are assigned to a calendar year using `claim_end_date` (raw
`CLM_THRU_DT`), per the DUG's own footnote to Table 1: *"Year is determined
by the claim through date."*

```{r reconcile-build}
ip_by_bene_year <- inpatient_clean %>%
  mutate(claim_year = year(claim_end_date)) %>%
  group_by(beneficiary_id, reference_year = claim_year) %>%
  summarise(
    claims_medicare_ip = sum(medicare_payment_amt + utilization_day_count * pass_through_per_diem_amt, na.rm = TRUE),
    claims_benres_ip = sum(beneficiary_deductible_amt + beneficiary_part_a_coinsurance_amt + beneficiary_blood_deductible_amt, na.rm = TRUE),
    claims_pppymt_ip = sum(primary_payer_paid_amt, na.rm = TRUE),
    .groups = "drop"
  )

op_by_bene_year <- outpatient_clean %>%
  mutate(claim_year = year(claim_end_date)) %>%
  group_by(beneficiary_id, reference_year = claim_year) %>%
  summarise(
    claims_medicare_op = sum(medicare_payment_amt, na.rm = TRUE),
    claims_benres_op = sum(beneficiary_blood_deductible_amt + beneficiary_part_b_deductible_amt + beneficiary_part_b_coinsurance_amt, na.rm = TRUE),
    claims_pppymt_op = sum(primary_payer_paid_amt, na.rm = TRUE),
    .groups = "drop"
  )

reconcile <- beneficiary_all %>%
  select(beneficiary_id, reference_year,
         annual_medicare_reimb_ip, annual_beneficiary_resp_ip, annual_primary_payer_reimb_ip,
         annual_medicare_reimb_op, annual_beneficiary_resp_op, annual_primary_payer_reimb_op) %>%
  left_join(ip_by_bene_year, by = c("beneficiary_id", "reference_year")) %>%
  left_join(op_by_bene_year, by = c("beneficiary_id", "reference_year")) %>%
  mutate(across(starts_with("claims_"), ~ replace_na(.x, 0)))
```

```{r reconcile-summary}
tibble(
  component = c("Inpatient: Medicare reimbursement", "Inpatient: beneficiary responsibility",
                "Inpatient: primary payer", "Outpatient: Medicare reimbursement",
                "Outpatient: beneficiary responsibility", "Outpatient: primary payer"),
  beneficiary_file_total = c(sum(reconcile$annual_medicare_reimb_ip), sum(reconcile$annual_beneficiary_resp_ip),
                             sum(reconcile$annual_primary_payer_reimb_ip), sum(reconcile$annual_medicare_reimb_op),
                             sum(reconcile$annual_beneficiary_resp_op), sum(reconcile$annual_primary_payer_reimb_op)),
  claims_derived_total = c(sum(reconcile$claims_medicare_ip), sum(reconcile$claims_benres_ip),
                           sum(reconcile$claims_pppymt_ip), sum(reconcile$claims_medicare_op),
                           sum(reconcile$claims_benres_op), sum(reconcile$claims_pppymt_op))
) %>%
  knitr::kable(digits = 2, format.args = list(big.mark = ","),
               caption = "Annual totals (Beneficiary file) vs. claims-derived totals, 2008-2010")
```

```{r reconcile-mismatch}
tol <- 0.005
tibble(
  component = c("Inpatient: Medicare reimbursement", "Inpatient: beneficiary responsibility",
                "Inpatient: primary payer", "Outpatient: Medicare reimbursement",
                "Outpatient: beneficiary responsibility", "Outpatient: primary payer"),
  beneficiary_years_differing = c(
    sum(abs(reconcile$annual_medicare_reimb_ip - reconcile$claims_medicare_ip) > tol, na.rm = TRUE),
    sum(abs(reconcile$annual_beneficiary_resp_ip - reconcile$claims_benres_ip) > tol, na.rm = TRUE),
    sum(abs(reconcile$annual_primary_payer_reimb_ip - reconcile$claims_pppymt_ip) > tol, na.rm = TRUE),
    sum(abs(reconcile$annual_medicare_reimb_op - reconcile$claims_medicare_op) > tol, na.rm = TRUE),
    sum(abs(reconcile$annual_beneficiary_resp_op - reconcile$claims_benres_op) > tol, na.rm = TRUE),
    sum(abs(reconcile$annual_primary_payer_reimb_op - reconcile$claims_pppymt_op) > tol, na.rm = TRUE)),
  beneficiary_years_na_in_beneficiary_file = c(
    sum(is.na(reconcile$annual_medicare_reimb_ip)), sum(is.na(reconcile$annual_beneficiary_resp_ip)),
    sum(is.na(reconcile$annual_primary_payer_reimb_ip)), sum(is.na(reconcile$annual_medicare_reimb_op)),
    sum(is.na(reconcile$annual_beneficiary_resp_op)), sum(is.na(reconcile$annual_primary_payer_reimb_op)))
) %>%
  knitr::kable(format.args = list(big.mark = ","),
               caption = "Beneficiary-years where the Beneficiary file value differs from the claims-derived value (tolerance of half a cent), and where it is NA")
```

### Why they match — a finding worth flagging

**Going in, the expectation (based on the DUG's general synthesis
methodology) was that these *wouldn't* reconcile.** Section 6 ("Methodology
and Limitation") of the Data User's Guide describes each synthetic
beneficiary starting from a real "seed," with *"the claims from the seed
beneficiary... replaced with claims from other donor beneficiary claims
sets"* — donor substitution that would normally decouple a beneficiary's
summary totals from whatever claims ended up attached to their synthetic
ID.

**The actual result above is the opposite: an exact match.** For all six
components, the Beneficiary file's annual total equals the claims-derived
total (e.g., Medicare reimbursement of \$`r format(round(sum(reconcile$annual_medicare_reimb_ip)), big.mark = ",")` for Inpatient and
\$`r format(round(sum(reconcile$annual_medicare_reimb_op)), big.mark = ",")` for Outpatient, to the dollar). The reason is in the Codebook's **Appendix 2 ("List of
Reimbursement Variables")** used to build the formulas in the table above:
it defines `MEDREIMB_IP`, `BENRES_IP`, `PPPYMT_IP`, etc. as literal sums
*over the claim-level fields*, not as an independently-donor-sourced
figure. That strongly suggests CMS computed the six annual columns tested
here (Inpatient and Outpatient; the three Carrier columns were not tested)
*after* finalizing the synthetic claims for a given release, as a rollup of
those exact synthetic claims — so they're internally consistent with this
extract's Inpatient/Outpatient files by construction, not by coincidence.

This is a narrower guarantee than it might look, though: the donor
substitution the DUG describes still applies to *other* Beneficiary fields
— demographics, and especially the chronic-condition flags behind Sections
3–4 — which aren't derived from the claims files the same way and carry no
such reconciliation guarantee. CMS's FAQ underscores that distinction
directly: *"much of the interdependence and co-variation among
variables have been altered... conditional correlations... would not be
reliable."* Read as: the dollar totals tie out; the relationships between
conditions, demographics, and cost don't necessarily.

Sources: CMS *DE-SynPUF Data User's Guide*, Section 6 ("Methodology and
Limitation"), `Docs/synpuf_dug.pdf`; CMS *DE-SynPUF Codebook*, Appendix 2
("List of Reimbursement Variables"), `Docs/de10_codebook.pdf`; CMS *DE-SynPUF
FAQ*, `Docs/de10_faq.pdf`.

## 6. Things to note

- The reconciliation in Section 5 came back an **exact match** for
  Inpatient and Outpatient — the opposite of what the DUG's general
  donor-substitution methodology predicted going in, and traceable instead
  to the Codebook's Appendix 2 defining these annual fields as literal sums
  over this release's claims. That guarantee is specific to the six annual
  reimbursement columns tested (Inpatient and Outpatient; the three Carrier
  columns were not tested), though — it says nothing about the chronic
  condition flags (Sections 3–4) or demographics, which follow the more
  general donor-substitution path the DUG describes and carry no such
  reconciliation guarantee.
- `cost_by_disease` (Section 4) double/triple-counts claims across comorbid
  conditions by design (see that section) — don't sum the claim-count
  columns (`n_*`) across rows expecting them to equal total claim counts.
- Treat the condition prevalence and correlations (Section 3) and cost-by-disease
  breakdown (Section 4) as a demonstration of the pipeline and methodology,
  not as clinical or actuarial fact — both rest on relationships CMS's own
  FAQ warns were altered by the disclosure-limitation synthesis.
- Extending the reconciliation to Carrier would require re-deriving
  `LINE_PRCSG_IND_CD`-filtered sums from the raw Carrier files, since that
  column was dropped during cleaning — not done here since it wasn't asked
  for, but the formulas are in Codebook Appendix 2 if needed later.
