---
title: "CMS Synthetic Data (DE-SynPUF) — Full Sample 1 Pull, Cleaning, Validation"
author: "Ann Raymond"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
options(scipen = 999)
```

## 1. Overview

This notebook pulls **all five DE-SynPUF file types for Sample 1** (2008-2010)
and cleans each one. Two things run throughout:

1. **Pipeline log** — every time a step drops/collapses/joins/binds rows or
   columns, it's logged with rows/columns *before* and *after*. The full log is brought together once, at the end (Section 10).
2. **Validation check** — every raw row count is checked against the
   official counts published in CMS's own **DE-SynPUF Data User's Guide**
   (mainly Table 2, p.5). 

Every column is also renamed from CMS's cryptic `ALL_CAPS` field codes to a
descriptive `snake_case` name — sourced from the variable labels in CMS's
**DE-SynPUF Codebook** (`Docs/de10_codebook.pdf`), not guessed. Variables that are fully empty and not
relevant are dropped outright, and certain variables are recoded from
numeric codes to their actual categories using the codebook's 
tables, not left as opaque integers. The Inpatient, Outpatient, and Carrier
claims files also get a `total_claim_amt` (Medicare-paid + primary-payer-paid +
beneficiary-responsibility, using CMS's documented formula — see Section 5
for why Inpatient's needs a per-diem term the others don't), and the
Beneficiary file gets the equivalent `total_annual_amt`.

One data-quality fix lives in Sections 5 and 6: `SEGMENT = 2` rows are dropped
from Inpatient and Outpatient before any amount is calculated.

This notebook stops at cleaning, validation, and saving — it does not do
any downstream analysis. Chronic-condition prevalence and correlations,
claim-cost-by-disease breakdowns, and a reconciliation check of the annual vs. claim-level totals
all live in the companion notebook, **`cms_synpuf_analysis.Rmd`**, which
reads the cleaned CSVs this one produces.

Files pulled (Sample 1, 2008-2010):

| File type | Variables (raw) | Records (CMS, Sample 1) |
|---|---:|---:|
| Beneficiary Summary (2008/2009/2010) | 32 | 116,352 / 114,538 / 112,754 |
| Inpatient Claims | 81 | 66,773 |
| Outpatient Claims | 76 | 790,790 |
| Carrier Claims (segments A + B) | 142 | 2,370,667 + 2,370,668 = 4,741,335 |
| Prescription Drug Events (PDE) | 8 | 5,552,421 |

Source: [CMS DE-SynPUF download page](https://www.cms.gov/data-research/statistics-trends-and-reports/medicare-claims-synthetic-public-use-files/cms-2008-2010-data-entrepreneurs-synthetic-public-use-file-de-synpuf/de10-sample-1) ·
Documented counts: CMS *Data User's Guide* ("synpuf_dug.pdf"), Table 2 and summary tables for each file.
Variable labels and category codes: CMS *DE-SynPUF Codebook* ("de10_codebook.pdf").

## 2. Environment setup

```{r packages}
required_pkgs <- c("tidyverse", "lubridate", "janitor", "data.table")

installed <- rownames(installed.packages())
missing <- setdiff(required_pkgs, installed)
if (length(missing) > 0) {
  install.packages(missing, repos = "https://cloud.r-project.org")
}

library(tidyverse)
library(lubridate)
library(janitor)
library(data.table)
```

Carrier Claims and PDE are large (2.4M and 5.6M rows; the two carrier
segments are ~1.2GB each as CSV). Those are read with `data.table::fread()`
instead of `readr::read_csv()` — it's markedly lighter on memory for
wide/large files, which matters since each carrier segment is read, reduced,
and discarded (`rm()` + `gc()`) one at a time rather than held in memory
together. Beneficiary and Inpatient stay on `readr` since they're small.

### Pipeline log & validation check

`log_step()` and `validate_count()` provide a complete picture of before/after changes and checks at the end.
`show_last_change()` and `show_last_validation()` show the changes and checks, right where that step happens.

```{r ledger-infra}
pipeline_log <- tibble(
  dataset = character(), step = character(),
  rows_before = double(), rows_after = double(),
  cols_before = double(), cols_after = double(),
  note = character()
)

log_step <- function(dataset, step, rows_before, rows_after,
                      cols_before = NA, cols_after = NA, note = "") {
  pipeline_log <<- bind_rows(pipeline_log, tibble(
    dataset, step, rows_before, rows_after, cols_before, cols_after, note
  ))
  invisible(pipeline_log)
}

show_last_change <- function(n = 1) {
  tail(pipeline_log, n) %>%
    knitr::kable(format.args = list(big.mark = ","), caption = "Change to this data file")
}

validation_log <- tibble(file = character(), cms_documented = double(), actual = double())

validate_count <- function(file, documented, actual) {
  validation_log <<- bind_rows(validation_log, tibble(file, cms_documented = documented, actual = actual))
  invisible(validation_log)
}

show_last_validation <- function(n = 1) {
  tail(validation_log, n) %>%
    mutate(match = cms_documented == actual, diff = actual - cms_documented) %>%
    knitr::kable(format.args = list(big.mark = ","), caption = "Validation vs. CMS Data User's Guide, Sample 1")
}

# CMS 2008-2010 DE-SynPUF Data User's Guide, Table 2 (p.5), Sample 1 row
CMS_SAMPLE1 <- list(
  bene_2008 = 116352, bene_2009 = 114538, bene_2010 = 112754,
  ip = 66773, op = 790790,
  carrier_a = 2370667, carrier_b = 2370668, carrier_ab = 4741335,
  pde = 5552421
)
```

### Category code tables (from the Codebook)

`BENE_RACE_CD` and `SP_STATE_CODE` variables contain bare numbers. Both are
lookup codes CMS documents explicitly in the Codebook's per-variable
"Categories" tables (Pgs. 14 and 16-17 respectively), reproduced here for validation checks:

```{r code-tables}
# de10_codebook.pdf, "BENE_RACE_CD" — code 4 does not exist; "Other" seems to be a catch-all for
# Unknown/Other/Asian/Hispanic-miscoded/North American Native.
RACE_CODES <- c("1"="White", "2"="Black", "3"="Other", "5"="Hispanic")

# de10_codebook.pdf, "SP_STATE_CODE" — code 54 is the catch-all "Others"
# category CMS uses for small states/territories; 40 and 48 are not used.
STATE_CODES <- c(
  "01"="AL","02"="AK","03"="AZ","04"="AR","05"="CA","06"="CO","07"="CT","08"="DE","09"="DC",
  "10"="FL","11"="GA","12"="HI","13"="ID","14"="IL","15"="IN","16"="IA","17"="KS","18"="KY",
  "19"="LA","20"="ME","21"="MD","22"="MA","23"="MI","24"="MN","25"="MS","26"="MO","27"="MT",
  "28"="NE","29"="NV","30"="NH","31"="NJ","32"="NM","33"="NY","34"="NC","35"="ND","36"="OH",
  "37"="OK","38"="OR","39"="PA","41"="RI","42"="SC","43"="SD","44"="TN","45"="TX","46"="UT",
  "47"="VT","49"="VA","50"="WA","51"="WV","52"="WI","53"="WY","54"="Other"
)
```

## 3. Pull the data

Zips were downloaded from CMS's public download URLs into `Data/Raw/` and
extracted into `Data/Extracted/`. Re-running this chunk is idempotent — it
skips files that already exist. Paths are relative to this file's location
(`Markdown files/`), one level up into the project's `Data/` folder.

```{r download, eval=FALSE}
raw_dir <- "../Data/Raw"
extracted_dir <- "../Data/Extracted"
dir.create(raw_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(extracted_dir, recursive = TRUE, showWarnings = FALSE)

files <- list(
  "de1_0_2008_beneficiary_summary_file_sample_1.zip" =
    "https://www.cms.gov/research-statistics-data-and-systems/downloadable-public-use-files/synpufs/downloads/de1_0_2008_beneficiary_summary_file_sample_1.zip",
  "de1_0_2009_beneficiary_summary_file_sample_1.zip" =
    "https://www.cms.gov/research-statistics-data-and-systems/downloadable-public-use-files/synpufs/downloads/de1_0_2009_beneficiary_summary_file_sample_1.zip",
  "de1_0_2010_beneficiary_summary_file_sample_1.zip" =
    "https://www.cms.gov/research-statistics-data-and-systems/statistics-trends-and-reports/synpufs/downloads/de1_0_2010_beneficiary_summary_file_sample_1.zip",
  "de1_0_2008_to_2010_inpatient_claims_sample_1.zip" =
    "https://www.cms.gov/research-statistics-data-and-systems/downloadable-public-use-files/synpufs/downloads/de1_0_2008_to_2010_inpatient_claims_sample_1.zip",
  "de1_0_2008_to_2010_outpatient_claims_sample_1.zip" =
    "https://www.cms.gov/research-statistics-data-and-systems/downloadable-public-use-files/synpufs/downloads/de1_0_2008_to_2010_outpatient_claims_sample_1.zip",
  "DE1_0_2008_to_2010_Carrier_Claims_Sample_1A.zip" =
    "http://downloads.cms.gov/files/DE1_0_2008_to_2010_Carrier_Claims_Sample_1A.zip",
  "DE1_0_2008_to_2010_Carrier_Claims_Sample_1B.zip" =
    "http://downloads.cms.gov/files/DE1_0_2008_to_2010_Carrier_Claims_Sample_1B.zip",
  "DE1_0_2008_to_2010_Prescription_Drug_Events_Sample_1.zip" =
    "http://downloads.cms.gov/files/DE1_0_2008_to_2010_Prescription_Drug_Events_Sample_1.zip"
)

for (fname in names(files)) {
  dest <- file.path(raw_dir, fname)
  if (!file.exists(dest)) {
    download.file(files[[fname]], destfile = dest, mode = "wb")
  }
  unzip(dest, exdir = extracted_dir, overwrite = TRUE)
}

list.files(extracted_dir)
```

## 4. Beneficiary Summary Files

```{r bene-load}
bene_2008_raw <- read_csv("../Data/Extracted/DE1_0_2008_Beneficiary_Summary_File_Sample_1.csv",
                           col_types = cols(.default = "c"))
bene_2009_raw <- read_csv("../Data/Extracted/DE1_0_2009_Beneficiary_Summary_File_Sample_1.csv",
                           col_types = cols(.default = "c"))
bene_2010_raw <- read_csv("../Data/Extracted/DE1_0_2010_Beneficiary_Summary_File_Sample_1.csv",
                           col_types = cols(.default = "c"))

validate_count("Beneficiary Summary 2008", CMS_SAMPLE1$bene_2008, nrow(bene_2008_raw))
validate_count("Beneficiary Summary 2009", CMS_SAMPLE1$bene_2009, nrow(bene_2009_raw))
validate_count("Beneficiary Summary 2010", CMS_SAMPLE1$bene_2010, nrow(bene_2010_raw))
show_last_validation(3)
```

All three years read in as character to start — the files mix dates, flags
coded `1`/`2`, and dollar amounts, so recoding happens explicitly below rather
than relying on `readr`'s guesses. None of the 32 raw columns are fully
empty in any of the three years, so the only column *dropped* here is
`BENE_COUNTY_CD` (by request, not because it's empty). Columns are renamed
to their descriptive, Codebook-sourced names *first* — then everything else
(type casts, recodes, the two new columns) is built on those final names.

```{r bene-clean}
clean_beneficiary <- function(df, year) {
  df %>%
    clean_names() %>%
    select(-bene_county_cd) %>%
    rename(
      beneficiary_id = desynpuf_id,
      birth_date = bene_birth_dt,
      death_date = bene_death_dt,
      sex = bene_sex_ident_cd,
      race = bene_race_cd,
      esrd_indicator = bene_esrd_ind,
      state = sp_state_code,
      part_a_coverage_months = bene_hi_cvrage_tot_mons,
      part_b_coverage_months = bene_smi_cvrage_tot_mons,
      hmo_coverage_months = bene_hmo_cvrage_tot_mons,
      part_d_coverage_months = plan_cvrg_mos_num,
      chronic_alzheimer = sp_alzhdmta,
      chronic_heart_failure = sp_chf,
      chronic_kidney_disease = sp_chrnkidn,
      chronic_cancer = sp_cncr,
      chronic_copd = sp_copd,
      chronic_depression = sp_depressn,
      chronic_diabetes = sp_diabetes,
      chronic_ischemic_heart_disease = sp_ischmcht,
      chronic_osteoporosis = sp_osteoprs,
      chronic_rheumatoid_or_osteoarthritis = sp_ra_oa,
      chronic_stroke_tia = sp_strketia,
      annual_medicare_reimb_ip = medreimb_ip,
      annual_beneficiary_resp_ip = benres_ip,
      annual_primary_payer_reimb_ip = pppymt_ip,
      annual_medicare_reimb_op = medreimb_op,
      annual_beneficiary_resp_op = benres_op,
      annual_primary_payer_reimb_op = pppymt_op,
      annual_medicare_reimb_car = medreimb_car,
      annual_beneficiary_resp_car = benres_car,
      annual_primary_payer_reimb_car = pppymt_car
    ) %>%
    mutate(
      birth_date = ymd(birth_date),
      death_date = ymd(death_date),
      across(
        c(chronic_alzheimer, chronic_heart_failure, chronic_kidney_disease, chronic_cancer,
          chronic_copd, chronic_depression, chronic_diabetes, chronic_ischemic_heart_disease,
          chronic_osteoporosis, chronic_rheumatoid_or_osteoarthritis, chronic_stroke_tia),
        ~ .x == "1"
      ),
      # number of the 11 CMS-flagged chronic conditions this beneficiary has
      # in this year — a simple row-sum now that the flags are boolean
      chronic_condition_count = rowSums(across(c(
        chronic_alzheimer, chronic_heart_failure, chronic_kidney_disease, chronic_cancer,
        chronic_copd, chronic_depression, chronic_diabetes, chronic_ischemic_heart_disease,
        chronic_osteoporosis, chronic_rheumatoid_or_osteoarthritis, chronic_stroke_tia
      ))),
      sex = recode(sex, "1" = "male", "2" = "female"),
      esrd_indicator = esrd_indicator == "Y",
      race = unname(RACE_CODES[race]),
      state = unname(STATE_CODES[state]),
      across(
        c(annual_medicare_reimb_ip, annual_beneficiary_resp_ip, annual_primary_payer_reimb_ip,
          annual_medicare_reimb_op, annual_beneficiary_resp_op, annual_primary_payer_reimb_op,
          annual_medicare_reimb_car, annual_beneficiary_resp_car, annual_primary_payer_reimb_car),
        as.numeric
      ),
      # total annual amount = Medicare-paid + primary-payer-paid + beneficiary
      # responsibility, summed across all three claim types (IP/OP/Carrier) —
      # the same three-part total the claims files carry, rolled up to a
      # single beneficiary-year figure
      total_annual_amt = annual_medicare_reimb_ip + annual_beneficiary_resp_ip + annual_primary_payer_reimb_ip +
        annual_medicare_reimb_op + annual_beneficiary_resp_op + annual_primary_payer_reimb_op +
        annual_medicare_reimb_car + annual_beneficiary_resp_car + annual_primary_payer_reimb_car,
      reference_year = year
    )
}

bene_2008 <- clean_beneficiary(bene_2008_raw, 2008)
bene_2009 <- clean_beneficiary(bene_2009_raw, 2009)
bene_2010 <- clean_beneficiary(bene_2010_raw, 2010)

rows_each <- nrow(bene_2008) + nrow(bene_2009) + nrow(bene_2010)
beneficiary_all <- bind_rows(bene_2008, bene_2009, bene_2010) %>%
  relocate(reference_year, .after = beneficiary_id)

cat("NA values after recoding: race =", sum(is.na(beneficiary_all$race)),
    "| state =", sum(is.na(beneficiary_all$state)), "\n")

log_step("beneficiary", "rename all columns first; drop county; recode race and state; create 2 new variables total_annual_amt and chronic_condition_count; row-bind files across the 3 years",
         rows_each, nrow(beneficiary_all),
         ncol(bene_2008_raw), ncol(beneficiary_all),
         "dropped 1 col (bene_county_cd); added 3 cols (reference_year + total_annual_amt + chronic_condition_count) = net +2 columns vs. raw")
show_last_change(1)
```

### Sanity checks

```{r bene-checks}
beneficiary_all %>% count(reference_year, name = "n_beneficiaries") %>% knitr::kable()

dup_check <- beneficiary_all %>% count(beneficiary_id, reference_year) %>% filter(n > 1) %>% nrow()
cat("Duplicate (id, year) pairs:", dup_check, "\n")

beneficiary_all %>% count(race, name = "n") %>% knitr::kable(caption = "Race, recoded")
beneficiary_all %>% count(state, name = "n") %>% arrange(desc(n)) %>% head(5) %>%
  knitr::kable(caption = "State, recoded (top 5)")

beneficiary_all %>%
  summarise(mean_total_annual_amt = mean(total_annual_amt, na.rm = TRUE),
            median_total_annual_amt = median(total_annual_amt, na.rm = TRUE),
            max_total_annual_amt = max(total_annual_amt, na.rm = TRUE)) %>%
  knitr::kable(digits = 2,
               caption = "total_annual_amt (USD) = Medicare-paid + primary-payer-paid + beneficiary responsibility, summed across IP/OP/Carrier")

beneficiary_all %>% count(chronic_condition_count, name = "n_beneficiary_years") %>%
  knitr::kable(caption = "chronic_condition_count distribution (0-11 conditions)")
```

## 5. Inpatient Claims

```{r ip-load}
ip_raw <- fread("../Data/Extracted/DE1_0_2008_to_2010_Inpatient_Claims_Sample_1.csv", colClasses = "character")
validate_count("Inpatient Claims (2008-2010)", CMS_SAMPLE1$ip, nrow(ip_raw))
show_last_validation(1)
```

```{r ip-empty-check}
ip_na_rate <- sapply(names(ip_raw), function(cn) mean(is.na(ip_raw[[cn]]) | ip_raw[[cn]] == ""))
ip_empty_cols <- names(ip_na_rate)[ip_na_rate == 1]
cat("Fully-empty columns in this sample:", length(ip_empty_cols), "of", ncol(ip_raw), "\n")
cat(paste(ip_empty_cols, collapse = ", "), "\n")
```

Every `HCPCS_CD_*` slot (all 45 of them) is 100% empty in this sample —
inpatient claims are DRG-priced and don't carry line-level HCPCS codes the
way outpatient/carrier claims do, so this isn't a data problem, it's
expected structure. They're dropped here, before anything else.

### Renaming, dropping incomplete rows, and calculating `total_claim_amt`

`SEGMENT = 2` records are incomplete or inconsistent (no diagnosis codes,
admitting diagnosis, or utilization days), and dropping them makes the claim
totals match the Beneficiary file's annual totals (confirmed below), so
they're excluded.

Columns are renamed to their descriptive names next, `SEGMENT = 2` rows are
then dropped (and the now-constant `claim_line_segment` column removed with
them), and the three beneficiary-liability columns plus
`pass_through_per_diem_amt` — previously left as raw strings — are cast to
numeric so `total_claim_amt` can sum them. Per the Codebook's Appendix 2
("List of Reimbursement Variables"), the Medicare-paid component of an
inpatient claim is `CLM_PMT_AMT + CLM_UTLZTN_DAY_CNT * CLM_PASS_THRU_PER_DIEM_AMT`
— not `CLM_PMT_AMT` alone — so `total_claim_amt` below includes that
per-diem term. `beneficiary_deductible_amt` is blank on about 3% of
claims. A blank means no deductible applied to the claim (the Part A
deductible is charged once per benefit period, and CMS fills the field only
when a deductible value code is present), so it is treated as \$0 in
`total_claim_amt`. The column itself stays `NA`, so the reconciliation, which
sums each component separately, is unchanged.

```{r ip-clean}
ip_cols_before <- ncol(ip_raw)
ip_reduced <- ip_raw[, setdiff(names(ip_raw), ip_empty_cols), with = FALSE]

ip_renamed <- ip_reduced %>%
  as_tibble() %>%
  clean_names() %>%
  rename(
    beneficiary_id = desynpuf_id,
    claim_id = clm_id,
    claim_line_segment = segment,
    claim_start_date = clm_from_dt,
    claim_end_date = clm_thru_dt,
    provider_institution_id = prvdr_num,
    medicare_payment_amt = clm_pmt_amt,
    primary_payer_paid_amt = nch_prmry_pyr_clm_pd_amt,
    attending_physician_npi = at_physn_npi,
    operating_physician_npi = op_physn_npi,
    other_physician_npi = ot_physn_npi,
    admission_date = clm_admsn_dt,
    admitting_diagnosis_code = admtng_icd9_dgns_cd,
    pass_through_per_diem_amt = clm_pass_thru_per_diem_amt,
    beneficiary_deductible_amt = nch_bene_ip_ddctbl_amt,
    beneficiary_part_a_coinsurance_amt = nch_bene_pta_coinsrnc_lblty_am,
    beneficiary_blood_deductible_amt = nch_bene_blood_ddctbl_lblty_am,
    utilization_day_count = clm_utlztn_day_cnt,
    discharge_date = nch_bene_dschrg_dt,
    drg_code = clm_drg_cd
  ) %>%
  rename_with(~ str_replace(., "^icd9_dgns_cd_", "diagnosis_code_"), starts_with("icd9_dgns_cd_")) %>%
  rename_with(~ str_replace(., "^icd9_prcdr_cd_", "procedure_code_"), starts_with("icd9_prcdr_cd_"))

ip_rows_before_segment_drop <- nrow(ip_renamed)
ip_cols_before_segment_drop <- ncol(ip_renamed)

ip_complete <- ip_renamed %>%
  filter(claim_line_segment == "1") %>%
  select(-claim_line_segment)

log_step("inpatient", "drop incomplete or inconsistent SEGMENT = 2 records (no diagnosis, admitting diagnosis, or utilization days); drop now-constant claim_line_segment column",
         ip_rows_before_segment_drop, nrow(ip_complete), ip_cols_before_segment_drop, ncol(ip_complete),
         sprintf("dropped %d rows; dropped 1 col (claim_line_segment)",
                 ip_rows_before_segment_drop - nrow(ip_complete)))
show_last_change(1)

inpatient_clean <- ip_complete %>%
  mutate(
    claim_start_date = ymd(claim_start_date),
    claim_end_date = ymd(claim_end_date),
    admission_date = ymd(admission_date),
    discharge_date = ymd(discharge_date),
    medicare_payment_amt = as.numeric(medicare_payment_amt),
    primary_payer_paid_amt = as.numeric(primary_payer_paid_amt),
    beneficiary_deductible_amt = as.numeric(beneficiary_deductible_amt),
    beneficiary_part_a_coinsurance_amt = as.numeric(beneficiary_part_a_coinsurance_amt),
    beneficiary_blood_deductible_amt = as.numeric(beneficiary_blood_deductible_amt),
    pass_through_per_diem_amt = as.numeric(pass_through_per_diem_amt),
    utilization_day_count = as.integer(utilization_day_count),
    length_of_stay = as.integer(discharge_date - admission_date),
    # total amount = (Medicare-paid + per-diem pass-through, per Codebook
    # Appendix 2) + primary-payer-paid + beneficiary liability (deductible +
    # Part A coinsurance + blood deductible). A blank beneficiary_deductible_amt
    # means no deductible applied to the claim (the Part A deductible is charged
    # once per benefit period, and CMS fills the field only when a deductible
    # value code is present), so it is treated as $0 here via coalesce(). The
    # column itself stays NA, so the reconciliation, which sums each component
    # separately, is unchanged.
    total_claim_amt = medicare_payment_amt + utilization_day_count * pass_through_per_diem_amt +
      primary_payer_paid_amt +
      coalesce(beneficiary_deductible_amt, 0) + beneficiary_part_a_coinsurance_amt + beneficiary_blood_deductible_amt
  )

log_step("inpatient", "cast types; add length_of_stay and total_claim_amt (incl. per-diem pass-through term)",
         nrow(ip_complete), nrow(inpatient_clean), ncol(ip_complete), ncol(inpatient_clean),
         "added 2 cols (length_of_stay, total_claim_amt); no rows affected")
show_last_change(1)
```

```{r ip-summary}
cat("Claims with blank beneficiary_deductible_amt (treated as $0 in total_claim_amt):",
    sum(is.na(inpatient_clean$beneficiary_deductible_amt)), "of", nrow(inpatient_clean), "\n")
cat("Claims with NA total_claim_amt:",
    sum(is.na(inpatient_clean$total_claim_amt)), "of", nrow(inpatient_clean), "\n")

inpatient_clean %>%
  summarise(n_claims = n(), n_beneficiaries = n_distinct(beneficiary_id),
            total_payment_before_per_diem_usd = sum(medicare_payment_amt, na.rm = TRUE),
            total_claim_amt_usd = sum(total_claim_amt, na.rm = TRUE),
            mean_length_of_stay_days = mean(length_of_stay, na.rm = TRUE),
            median_length_of_stay_days = median(length_of_stay, na.rm = TRUE)) %>%
  knitr::kable(digits = 2)

tibble(
  component = c("Medicare-paid", "Beneficiary responsibility", "Primary payer paid"),
  this_file_usd = c(
    sum(inpatient_clean$medicare_payment_amt + inpatient_clean$utilization_day_count * inpatient_clean$pass_through_per_diem_amt, na.rm = TRUE),
    sum(inpatient_clean$beneficiary_deductible_amt + inpatient_clean$beneficiary_part_a_coinsurance_amt + inpatient_clean$beneficiary_blood_deductible_amt, na.rm = TRUE),
    sum(inpatient_clean$primary_payer_paid_amt, na.rm = TRUE)
  ),
  beneficiary_file_usd = c(
    sum(beneficiary_all$annual_medicare_reimb_ip, na.rm = TRUE),
    sum(beneficiary_all$annual_beneficiary_resp_ip, na.rm = TRUE),
    sum(beneficiary_all$annual_primary_payer_reimb_ip, na.rm = TRUE)
  )
) %>%
  mutate(match = this_file_usd == beneficiary_file_usd) %>%
  knitr::kable(digits = 2, caption = "Confirming the SEGMENT = 2 drop against the Beneficiary file")
```

### Referential check against beneficiaries

```{r ip-join-check}
ip_orphans <- inpatient_clean %>% distinct(beneficiary_id) %>% anti_join(beneficiary_all, by = "beneficiary_id") %>% nrow()
cat("Inpatient beneficiaries with no match in the beneficiary file:", ip_orphans, "\n")
```

## 6. Outpatient Claims

```{r op-load}
op_raw <- fread("../Data/Extracted/DE1_0_2008_to_2010_Outpatient_Claims_Sample_1.csv", colClasses = "character")
validate_count("Outpatient Claims (2008-2010)", CMS_SAMPLE1$op, nrow(op_raw))
show_last_validation(1)
```

Unlike inpatient, outpatient claims *do* use HCPCS codes for line-level
procedures — `HCPCS_CD_1` is populated on 82% of claims, and fill rates
gradually decline to `HCPCS_CD_44`. Only `HCPCS_CD_45` (the very last
HCPCS variable) is fully empty — checked the same way as inpatient, not assumed:

```{r op-empty-check}
op_na_rate <- sapply(names(op_raw), function(cn) mean(is.na(op_raw[[cn]]) | op_raw[[cn]] == ""))
op_empty_cols <- names(op_na_rate)[op_na_rate == 1]
cat("Fully-empty columns:", length(op_empty_cols), "of", ncol(op_raw), "—", paste(op_empty_cols, collapse=", "), "\n")
```

### Renaming, dropping incomplete rows, and calculating `total_claim_amt`

As with Inpatient, `SEGMENT = 2` records are incomplete or inconsistent
(blank claim dates, often no diagnosis) and are excluded. Columns are
renamed first, then those rows are dropped along with the now-constant
`claim_line_segment` column, before the
type casts and `total_claim_amt` (Medicare-paid + primary-payer-paid +
beneficiary liability — Outpatient's formula has no per-diem term, per
Codebook Appendix 2).

```{r op-clean}
op_cols_before <- ncol(op_raw)
op_reduced <- op_raw[, setdiff(names(op_raw), op_empty_cols), with = FALSE]

op_renamed <- op_reduced %>%
  as_tibble() %>%
  clean_names() %>%
  rename(
    beneficiary_id = desynpuf_id,
    claim_id = clm_id,
    claim_line_segment = segment,
    claim_start_date = clm_from_dt,
    claim_end_date = clm_thru_dt,
    provider_institution_id = prvdr_num,
    medicare_payment_amt = clm_pmt_amt,
    primary_payer_paid_amt = nch_prmry_pyr_clm_pd_amt,
    attending_physician_npi = at_physn_npi,
    operating_physician_npi = op_physn_npi,
    other_physician_npi = ot_physn_npi,
    beneficiary_blood_deductible_amt = nch_bene_blood_ddctbl_lblty_am,
    beneficiary_part_b_deductible_amt = nch_bene_ptb_ddctbl_amt,
    beneficiary_part_b_coinsurance_amt = nch_bene_ptb_coinsrnc_amt,
    admitting_diagnosis_code = admtng_icd9_dgns_cd
  ) %>%
  rename_with(~ str_replace(., "^icd9_dgns_cd_", "diagnosis_code_"), starts_with("icd9_dgns_cd_")) %>%
  rename_with(~ str_replace(., "^icd9_prcdr_cd_", "procedure_code_"), starts_with("icd9_prcdr_cd_")) %>%
  rename_with(~ str_replace(., "^hcpcs_cd_", "procedure_hcpcs_code_"), starts_with("hcpcs_cd_"))

op_rows_before_segment_drop <- nrow(op_renamed)
op_cols_before_segment_drop <- ncol(op_renamed)

op_complete <- op_renamed %>%
  filter(claim_line_segment == "1") %>%
  select(-claim_line_segment)

log_step("outpatient", "drop incomplete or inconsistent SEGMENT = 2 records (blank claim dates, often no diagnosis); drop now-constant claim_line_segment column",
         op_rows_before_segment_drop, nrow(op_complete), op_cols_before_segment_drop, ncol(op_complete),
         sprintf("dropped %d rows; dropped 1 col (claim_line_segment)",
                 op_rows_before_segment_drop - nrow(op_complete)))
show_last_change(1)

outpatient_clean <- op_complete %>%
  mutate(
    claim_start_date = ymd(claim_start_date),
    claim_end_date = ymd(claim_end_date),
    medicare_payment_amt = as.numeric(medicare_payment_amt),
    primary_payer_paid_amt = as.numeric(primary_payer_paid_amt),
    beneficiary_blood_deductible_amt = as.numeric(beneficiary_blood_deductible_amt),
    beneficiary_part_b_deductible_amt = as.numeric(beneficiary_part_b_deductible_amt),
    beneficiary_part_b_coinsurance_amt = as.numeric(beneficiary_part_b_coinsurance_amt),
    # total amount = Medicare-paid + primary-payer-paid + beneficiary liability
    # (blood deductible + Part B deductible + Part B coinsurance)
    total_claim_amt = medicare_payment_amt + primary_payer_paid_amt +
      beneficiary_blood_deductible_amt + beneficiary_part_b_deductible_amt + beneficiary_part_b_coinsurance_amt
  )

log_step("outpatient", "cast types; add total_claim_amt", nrow(op_complete), nrow(outpatient_clean),
         ncol(op_complete), ncol(outpatient_clean), "added 1 col (total_claim_amt); no rows affected")
show_last_change(1)
rm(op_raw, op_reduced, op_renamed, op_complete); invisible(gc(verbose = FALSE))
```

```{r op-summary}
outpatient_clean %>%
  summarise(n_claims = n(), n_beneficiaries = n_distinct(beneficiary_id),
            total_medicare_paid_usd = sum(medicare_payment_amt, na.rm = TRUE),
            total_claim_amt_usd = sum(total_claim_amt, na.rm = TRUE)) %>%
  knitr::kable(digits = 2)

tibble(
  component = c("Medicare-paid", "Beneficiary responsibility", "Primary payer paid"),
  this_file_usd = c(
    sum(outpatient_clean$medicare_payment_amt, na.rm = TRUE),
    sum(outpatient_clean$beneficiary_blood_deductible_amt + outpatient_clean$beneficiary_part_b_deductible_amt + outpatient_clean$beneficiary_part_b_coinsurance_amt, na.rm = TRUE),
    sum(outpatient_clean$primary_payer_paid_amt, na.rm = TRUE)
  ),
  beneficiary_file_usd = c(
    sum(beneficiary_all$annual_medicare_reimb_op, na.rm = TRUE),
    sum(beneficiary_all$annual_beneficiary_resp_op, na.rm = TRUE),
    sum(beneficiary_all$annual_primary_payer_reimb_op, na.rm = TRUE)
  )
) %>%
  mutate(match = this_file_usd == beneficiary_file_usd) %>%
  knitr::kable(digits = 2, caption = "Confirming the SEGMENT = 2 drop against the Beneficiary file")
```

### Referential check against beneficiaries

```{r op-join-check}
op_orphans <- outpatient_clean %>% distinct(beneficiary_id) %>% anti_join(beneficiary_all, by = "beneficiary_id") %>% nrow()
cat("Outpatient beneficiaries with no match in the beneficiary file:", op_orphans, "\n")
```

## 7. Carrier Claims (Segments A + B)

Carrier claims are the largest file type by far: 142 columns, ~2.37M rows
*per segment* (CMS split each subsample into A/B purely for file-size
reasons — it's one logical file; this "segment" is unrelated to the
`SEGMENT` field handled in Sections 5-6, which Carrier doesn't have).
Each segment is ~1.2GB as CSV, so both segments are never held in memory at
once: each is read, reduced to a claim-level summary, and discarded
(`rm()` + `gc()`) before the next one loads.

142 columns is mostly *repetition*: up to 13 service lines per claim, each
with its own physician NPI, tax number, HCPCS code, and dollar amounts for
Medicare payment, primary-payer payment, Part B deductible, coinsurance, and
allowed charge — plus a processing-indicator code and a line-level diagnosis
code. Rather than rename those 1:1, the cleaning step collapses them to a
per-claim summary, with the final column names assigned directly rather
than renamed afterward: line count, primary HCPCS code, and totals across
*all three* payer components CMS tracks at the line level (Medicare paid,
primary payer paid, and beneficiary liability). Per-line physician NPIs,
tax numbers, processing-indicator codes, and line-level diagnosis pointers
are still dropped — those don't roll up into a meaningful per-claim total.

```{r carrier-summarize-fn}
summarize_carrier <- function(dt) {
  hcpcs_cols      <- grep("^HCPCS_CD_", names(dt), value = TRUE)
  nch_cols        <- grep("^LINE_NCH_PMT_AMT_", names(dt), value = TRUE)
  coins_cols      <- grep("^LINE_COINSRNC_AMT_", names(dt), value = TRUE)
  alowd_cols      <- grep("^LINE_ALOWD_CHRG_AMT_", names(dt), value = TRUE)
  ptb_ddctbl_cols <- grep("^LINE_BENE_PTB_DDCTBL_AMT_", names(dt), value = TRUE)
  prmry_pyr_cols  <- grep("^LINE_BENE_PRMRY_PYR_PD_AMT_", names(dt), value = TRUE)
  dgns_cols       <- grep("^ICD9_DGNS_CD_", names(dt), value = TRUE)

  sum_cols <- function(cols) Reduce(`+`, lapply(dt[, ..cols], function(x) {
    x[is.na(x) | x == ""] <- "0"; as.numeric(x)
  }))

  dt[, service_line_count := Reduce(`+`, lapply(.SD, function(x) !is.na(x) & x != "")), .SDcols = hcpcs_cols]
  dt[, primary_procedure_code := apply(.SD, 1, function(x) { v <- x[!is.na(x) & x != ""]; if (length(v)) v[1] else NA_character_ }), .SDcols = hcpcs_cols]
  dt[, diagnosis_code_count := Reduce(`+`, lapply(.SD, function(x) !is.na(x) & x != "")), .SDcols = dgns_cols]
  dt[, total_medicare_payment_amt := sum_cols(nch_cols)]
  dt[, total_beneficiary_coinsurance_amt := sum_cols(coins_cols)]
  dt[, total_allowed_charge_amt := sum_cols(alowd_cols)]
  dt[, total_beneficiary_deductible_amt := sum_cols(ptb_ddctbl_cols)]
  dt[, total_primary_payer_paid_amt := sum_cols(prmry_pyr_cols)]

  keep <- c("DESYNPUF_ID", "CLM_ID", "CLM_FROM_DT", "CLM_THRU_DT", dgns_cols,
            "diagnosis_code_count", "service_line_count", "primary_procedure_code",
            "total_medicare_payment_amt", "total_beneficiary_coinsurance_amt", "total_allowed_charge_amt",
            "total_beneficiary_deductible_amt", "total_primary_payer_paid_amt")
  out <- dt[, ..keep]
  setnames(out,
           old = c("DESYNPUF_ID", "CLM_ID", "CLM_FROM_DT", "CLM_THRU_DT", dgns_cols),
           new = c("beneficiary_id", "claim_id", "claim_start_date", "claim_end_date",
                   str_replace(dgns_cols, "^ICD9_DGNS_CD_", "diagnosis_code_")))
  out
}
```

### Segment A

```{r carrier-a}
car_a_raw <- fread("../Data/Extracted/DE1_0_2008_to_2010_Carrier_Claims_Sample_1A.csv", colClasses = "character")
validate_count("Carrier Claims A (2008-2010)", CMS_SAMPLE1$carrier_a, nrow(car_a_raw))
show_last_validation(1)

car_a_cols_before <- ncol(car_a_raw)
car_a_clean <- summarize_carrier(car_a_raw)

log_step("carrier_a", "collapse 13 service-line slots to per-claim summary; assign final column names directly",
         nrow(car_a_raw), nrow(car_a_clean), car_a_cols_before, ncol(car_a_clean),
         "dropped per-line NPI/tax-num/processing-indicator/line-level diagnosis; kept claim-level diagnosis codes + line-item sums (now including primary-payer-paid and Part B deductible)")
show_last_change(1)

rm(car_a_raw); invisible(gc(verbose = FALSE))
```

### Segment B

```{r carrier-b}
car_b_raw <- fread("../Data/Extracted/DE1_0_2008_to_2010_Carrier_Claims_Sample_1B.csv", colClasses = "character")
validate_count("Carrier Claims B (2008-2010)", CMS_SAMPLE1$carrier_b, nrow(car_b_raw))
show_last_validation(1)

car_b_cols_before <- ncol(car_b_raw)
car_b_clean <- summarize_carrier(car_b_raw)

log_step("carrier_b", "collapse 13 service-line slots to per-claim summary; assign final column names directly",
         nrow(car_b_raw), nrow(car_b_clean), car_b_cols_before, ncol(car_b_clean),
         "dropped per-line NPI/tax-num/processing-indicator/line-level diagnosis; kept claim-level diagnosis codes + line-item sums (now including primary-payer-paid and Part B deductible)")
show_last_change(1)

rm(car_b_raw); invisible(gc(verbose = FALSE))
```

### Combine A + B

```{r carrier-combine}
carrier_rows_before <- nrow(car_a_clean) + nrow(car_b_clean)
carrier_cols_before_combine <- ncol(car_a_clean)

carrier_clean <- rbindlist(list(car_a_clean, car_b_clean)) %>%
  as_tibble() %>%
  mutate(
    claim_start_date = ymd(claim_start_date),
    claim_end_date = ymd(claim_end_date),
    # total amount = Medicare-paid + primary-payer-paid + beneficiary liability
    # (Part B deductible + coinsurance), all summed across service lines
    total_claim_amt = total_medicare_payment_amt + total_primary_payer_paid_amt +
      total_beneficiary_coinsurance_amt + total_beneficiary_deductible_amt
  )

log_step("carrier", "row-bind segments A and B; add total_claim_amt", carrier_rows_before, nrow(carrier_clean),
         carrier_cols_before_combine, ncol(carrier_clean),
         "union of the two carrier segments CMS split for file-size reasons (no rows lost); added 1 col (total_claim_amt)")
show_last_change(1)

validate_count("Carrier Claims A+B (2008-2010)", CMS_SAMPLE1$carrier_ab, nrow(carrier_clean))
show_last_validation(1)
rm(car_a_clean, car_b_clean); invisible(gc(verbose = FALSE))
```

### Referential check against beneficiaries

```{r carrier-join-check}
car_orphans <- carrier_clean %>% distinct(beneficiary_id) %>% anti_join(beneficiary_all, by = "beneficiary_id") %>% nrow()
cat("Carrier-claim beneficiaries with no match in the beneficiary file:", car_orphans, "\n")
```

## 8. Prescription Drug Events (PDE)

```{r pde-load}
pde_raw <- fread("../Data/Extracted/DE1_0_2008_to_2010_Prescription_Drug_Events_Sample_1.csv", colClasses = "character")
validate_count("Prescription Drug Events (2008-2010)", CMS_SAMPLE1$pde, nrow(pde_raw))
show_last_validation(1)
```

PDE is the simplest file: 8 variables, no repeated line-item structure, and
(checked the same way as the others) zero fully-empty columns — so this
step is renaming and type casts only, no drops, no new totals.

```{r pde-clean}
pde_clean <- pde_raw %>%
  as_tibble() %>%
  clean_names() %>%
  rename(
    beneficiary_id = desynpuf_id,
    prescription_event_id = pde_id,
    service_date = srvc_dt,
    product_service_id = prod_srvc_id,
    quantity_dispensed = qty_dspnsd_num,
    days_supply = days_suply_num,
    patient_pay_amt = ptnt_pay_amt,
    gross_drug_cost_amt = tot_rx_cst_amt
  ) %>%
  mutate(
    service_date = ymd(service_date),
    quantity_dispensed = as.numeric(quantity_dispensed),
    days_supply = as.integer(days_supply),
    patient_pay_amt = as.numeric(patient_pay_amt),
    gross_drug_cost_amt = as.numeric(gross_drug_cost_amt)
  )

log_step("pde", "rename all columns first, then cast types (no drops, no new totals requested)", nrow(pde_raw), nrow(pde_clean),
         ncol(pde_raw), ncol(pde_clean), "0 fully-empty columns found")
show_last_change(1)
rm(pde_raw); invisible(gc(verbose = FALSE))
```

### Referential check against beneficiaries

```{r pde-join-check}
pde_orphans <- pde_clean %>% distinct(beneficiary_id) %>% anti_join(beneficiary_all, by = "beneficiary_id") %>% nrow()
cat("PDE beneficiaries with no match in the beneficiary file:", pde_orphans, "\n")
```

## 9. Save cleaned outputs

```{r save}
dir.create("../Data/Clean", recursive = TRUE, showWarnings = FALSE)
write_csv(beneficiary_all, "../Data/Clean/beneficiary_2008_2010.csv")
write_csv(inpatient_clean, "../Data/Clean/inpatient_claims_sample_1.csv")
write_csv(outpatient_clean, "../Data/Clean/outpatient_claims_sample_1.csv")
write_csv(carrier_clean, "../Data/Clean/carrier_claims_sample_1.csv")
write_csv(pde_clean, "../Data/Clean/prescription_drug_events_sample_1.csv")

cat("Saved 5 cleaned files to Data/Clean/\n")
```

## 10. Master pipeline log

Every row/column drop, collapse, and bind from this run, in one place:

```{r master-ledger}
knitr::kable(pipeline_log, format.args = list(big.mark = ","), caption = "Master pipeline log")
```

## 11. Master validation vs. CMS Data User's Guide

Every raw file's row count against CMS's own documented Sample 1 counts
(Table 2, p.5 of the Data User's Guide) — `match` should read `TRUE` across
the board:

```{r master-validation}
validation_log %>%
  mutate(match = cms_documented == actual, diff = actual - cms_documented) %>%
  knitr::kable(format.args = list(big.mark = ","), caption = "Master validation")
```

Raw column counts were also checked against the summary tables for each file in the same
guide, which states 32 / 81 / 76 / 142 / 8 variables for Beneficiary /
Inpatient / Outpatient / Carrier / PDE respectively — every file's raw
`ncol()` above matched before any columns were dropped or renamed.

## 12. Summary of changes

- `SEGMENT = 2` rows are dropped from Inpatient and Outpatient *before*
  `total_claim_amt` is calculated (Sections 5-6). They are incomplete
  or inconsistent records, and with them dropped every IP/OP reimbursement
  component sums exactly to the Beneficiary file's annual total.
- `total_claim_amt` for Inpatient includes the per-diem pass-through term
  (`utilization_day_count * pass_through_per_diem_amt`) per the Codebook's
  Appendix 2 ("List of Reimbursement Variables"). Outpatient and Carrier's
  formulas don't have an equivalent term, per the same Appendix.
- `total_claim_amt` (IP/OP/Carrier) and `total_annual_amt` (Beneficiary) sum
  Medicare-paid + primary-payer-paid + beneficiary liability. For OP,
  claim totals equal the sum of the three reconciled components exactly.
  For IP, `total_claim_amt` exceeds that sum by the coinsurance and
  blood-deductible amounts on the blank-deductible claims, which the
  reconciliation's beneficiary-responsibility figure leaves out. For
  Carrier it's a claim-level sum of already-summed line-level
  components, so it inherits whatever rounding CMS's synthetic-data process
  introduced at the line level. `total_annual_amt` is the beneficiary-year
  rollup of the same three components across all three claim types
  (IP+OP+Carrier) — it is *not* directly comparable to a single claim's
  `total_claim_amt`. (`cms_synpuf_analysis.Rmd` has a reconciliation check
  of these totals against the claims files.)
- The carrier-claims collapse (142 → 21 columns) is a real simplification,
  not a data-quality fix — if per-line physician or processing-indicator
  detail turns out to matter for a later analysis, that needs a reshape to
  long format instead of this claim-level summary.
- `race` and `state` are recoded per the Codebook's own category tables;
  both carry the Codebook's own disclosure-treatment caveats (race in
  particular is flagged as imputed/coarsened — CMS says not to draw
  population inferences from it).
- All four referential checks (inpatient/outpatient/carrier/PDE against the
  beneficiary file) came back clean above; worth re-running if a different
  sample or year range is pulled later.
- If sharing this notebook, note that although the data is fully synthetic,
  it's still meant to structurally resemble real CMS PUFs — treat it as a
  stand-in for real claims data, not for external distribution as "real"
  Medicare statistics.
- Downstream analysis (chronic-condition prevalence and correlations, cost-by-disease
  breakdowns, and the reconciliation check mentioned above) lives in the
  companion notebook **`cms_synpuf_analysis.Rmd`**, which reads the five
  CSVs saved in Section 9 rather than re-running this pipeline.
